Modeling the Latent Space of Symbolic String Quartet Music

### 2021-06-06 papers/final/alex/paper.tex
data augmentation for vae model

%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,authorversion]{acmart}
%% NOTE that a single column version may be required for
%% submission and peer review. This can be done by changing
%% the \doucmentclass[...]{acmart} in this template to
%% \documentclass[manuscript,screen,review]{acmart}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

In order to accommodate polyphonic music, we each MIDI file is
converted into a modified pianoroll representation, an example of
which is visualized in Figure \ref{pianoroll}. In this modified
representation, each instrument track is a sequence of integers from
0-129 representing the MIDI note pitch value being played at the given
time step, where values 0-127 are pitches, 128 is a rest (silence) and
129 is sustain, indicating that the previously played pitch is held
continuously. MIDI data contains pressure values that indicate how
hard or soft each note is played; this information is discarded in our
preprocessing and all notes are represented with constant
pressure. Because chords (multiple notes being played simultaneously
by the same instrument) are relatively rare in string quartets, the
lowest note of each chord is taken and the rest discarded, to reduce
dimensionality. Because notes at the extreme low and high end are rare
and are out of range of common instruments, the note pitch values are
clipped to a narrower range and ordinal encoded. For the 36 string
quartets found in the MusicNet database, this results in 66 distinct
values, where 0-63 represent notes played by the violins, viola and
cello, 64 is rest and 65 is sustain. Our software package also
includes functions to convert this representation back into a Music21
Score object, which can then be written to a MIDI file.

@@

Data augmentation is also possible and we assess its impact on
the results--the literature suggests augmentation via pitch shifting
each training example up or down by up to six semitones, and
increasing or reducing the speed by up to 10\% in order to create
additional training examples and reduce overfitting
\cite{oore_this_2018}.

The multi-stream representation discussed in
\cite{kumar2019polyphonic} addresses the issue of extreme class
imbalance and sparsity in the pianoroll representation.

@@

The autoencoder reconstruction loss function is sparse categorical
cross-entropy, which is added to the Kullback-Leibler divergence of
the approximate posterior distribution from a multivariate Gaussian
distribution with mean 0 and standard deviation 1, resulting in the
(negative) evidence lower bound (ELBO) loss, which is a standard
variational autoencoder loss function. This loss function is applied
to each track independently and then summed across all four tracks
with equal weight.

@@

\begin{itemize}
  \tightlist
  \item batch size: \textbf{32}, 64, 96
  \item learning rate: 0.001, \textbf{0.0002}
  \item recurrent layer type: LSTM, GRU (made no difference)
  \item recurrent encoder layer directions: \textbf{unidirectional}, bidirectional
  \item recurrent layer cells: 64, \textbf{128}, 256
  \item latent code dimension: 128, 256, \textbf{512}
  \item embedding dimension: 4, \textbf{8}, 16
  \item dropout rate: 0.2, 0.4, \textbf{0.5}
\end{itemize}

@@

@@

The reconstructed music produced by the model sounds reminiscent of
the original inputs, preserving the key signature and general rhythmic
and melodic structure, albeit with noticeable differences.

A website with several samples of the original input MIDI tracks, the
autoencoded outputs, and interpolations, is presented at: TODO

TODO: learning curve plot

TODO: links to audio samples

>>>>

%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,authorversion]{acmart}
\usepackage{pgf}
\usepackage{pgfplots}
%% NOTE that a single column version may be required for
%% submission and peer review. This can be done by changing
%% the \doucmentclass[...]{acmart} in this template to
%% \documentclass[manuscript,screen,review]{acmart}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

In order to accommodate polyphonic music, we each MIDI file is
converted into a modified pianoroll representation, an example of
which is visualized in Figure \ref{pianoroll}. While the standard
pianoroll represents each time-step as a multi-hot encoded vector of
note values, because most notes are not being played at most timesteps
there is an extreme class imbalance and sparsity in the matrix. This
is reported in \cite{kumar2019polyphonic} to cause problems in model
fitting and metrics calculation, so they introduce an alternative
``multi-stream'' representation of the pianoroll, which we closely,
(but not exactly) emulate in this work.

In our modified representation, each instrument track is a sequence of
integers from 0-129 representing the MIDI note pitch value being
played at the given time step, where values 0-127 are pitches, 128 is
a rest (silence) and 129 is sustain, indicating that the previously
played pitch is held continuously. MIDI data contains pressure values
that indicate how hard or soft each note is played; this information
is discarded in our preprocessing and all notes are represented with
constant pressure. Because chords (multiple notes being played
simultaneously by the same instrument) are relatively rare in string
quartets, the lowest note of each chord is taken and the rest
discarded, to reduce dimensionality. Because notes at the extreme low
and high end are rare and are out of range of common instruments, the
note pitch values are clipped to a narrower range and ordinal
encoded. For the 36 string quartets found in the MusicNet database,
this results in 66 distinct values, where 0-63 represent notes played
by the violins, viola and cello, 64 is rest and 65 is sustain. Our
software package also includes functions to convert this
representation back into a Music21 Score object, which can then be
written to a MIDI file.

TODO: plot histogram of notes per instrument

@@

@@

The autoencoder reconstruction loss function is sparse categorical
cross-entropy, which is added to the Kullback-Leibler divergence of
the approximate posterior distribution from a multivariate Gaussian
distribution with mean 0 and standard deviation 1, resulting in the
(negative) evidence lower bound (ELBO) loss (Equation \ref{nelbo}),
which is a standard variational autoencoder loss function. This loss
function is applied to each track independently and then summed across
all four tracks with equal weight.

\begin{equation}
  \label{nelbo}
\mathbb{E}[log{p_\theta}(x|z)]-\mathbb{KL}(q_\lambda(z|x)||p(z))
\end{equation}

where $\lambda$ and $\theta$ represent the encoder and decoder
parameters respectively, while $q_\lambda(z|x)$ represents the
encoder, which approximates the true posterior probability $p(z|x)$,
and $p_\theta(x|z)$ represents the decoder, which parameterizes the
likelihood $p(x|z)$ \cite{roberts_hierarchical_2018}.

@@

\begin{itemize}
  \tightlist
  \item batch size: \textbf{32}, 64, 96
  \item learning rate: 0.001, \textbf{0.0002}
  \item recurrent layer type: LSTM, GRU (made no difference)
  \item recurrent encoder layer directions: \textbf{unidirectional}, bidirectional
  \item recurrent layer cells: 64, \textbf{128}, 256
  \item latent code dimension: 128, 256, \textbf{512}
  \item embedding dimension: 4, \textbf{8}, 16
  \item dropout rate: 0.2, 0.4, \textbf{0.5}
  \item time steps per beat (quarter note): \textbf{4}, 8
  \item beats per phrase: \textbf{4}, 8
\end{itemize}

@@

Data augmentation was also tested--\cite{oore_this_2018} suggests
augmentation via pitch shifting each training example up or down by up
to six semitones in order to create additional training examples and
reduce overfitting, so we followed that recommendation and used it to
double the size of the training dataset. As a result, we saw much
lower validation losses (compare Figure \ref{loss} and Figure
\ref{loss_augment} and a small but noticeable improvement in the
fidelity of the song reconstructions, suggesting that this method was
effective in reducing overfitting.

\begin{figure}[htbp]
    \begin{center}
        \scalebox{0.5}{\input{figures/loss.pgf}}
    \end{center}
    \caption{Training and validation loss for the best performing model - without data augmentation}
    \label{loss}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
        \scalebox{0.5}{\input{figures/loss_augment.pgf}}
    \end{center}
    \caption{Training and validation loss for the best performing model - with data augmentation}
    \label{loss_augment}
\end{figure}

@@

The reconstructed music produced by the model sounds reminiscent of
the original inputs, preserving the key signature and general rhythmic
and melodic structure, albeit with noticeable differences at the
individual note level.

A website with several samples of the original input MIDI tracks, the
autoencoded outputs, and interpolations, is presented at: \textbf{TODO}
